---
title: "acorn Vignette"
author: "Tychele N. Turner, Ph.D., Washington University School of Medicine"
date: "2023-04-05"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{acorn Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load the acorn Package
```{r}
library('acorn')
```

### Load in the Test DNV Data
readDNV = Reads in a de novo variant (DNV) file in the format of sample, 
chromosome, genomic position, reference allele, alternate allele, and 
then any optional columns. File must be tab-delimited and the file must 
have the data in the order listed above (i.e., sample is field 1, chromosome 
is field 2, genomic position is field 3, reference allele is field 4, and 
alternate allele is field 5. The file can either be a uncompressed file or 
can be a gz compressed file. Please note that the chromosome data should 
take the form with a "chr" at the beginning (e.g., chr1).

Returns back a loaded in version of the DNV file that can be assigned to 
an object.
```{r}
input <- readDNV(paste(path.package("acorn"),"/extdata/dnms_from_Ng_et_al_2022_Human_Mutation_paper.txt.gz",sep="")) 
head(input)
str(input)
```

### Extract an Individual
extractIndividual = Extracts the DNVs out of a dnvObject from a particular 
individual. Returns a DNV object containing only DNVs in the specified
individual.
```{r}
ind <- extractIndividual(input, "HG01928")
head(ind)
nrow(ind)
table(ind[,1])
```

### Extract Multiple Individuals
```{r}
ind <- extractIndividual(input, c("HG01928", "HG03915"))
head(ind)
nrow(ind)
table(ind[,1])
```

### extract SNVs
extractSNVs = Extracts single-nucleotide variants (SNVs) out from a DNV object
generated using the readDNV function. Returns a DNV object containing only SNVs.
```{r}
snvs <- extractSNVs(input)
nrow(snvs)
```

### Extract INDELs Only
extractINDELs = Extracts small insertions/deletions (INDELs) out from a DNV 
object generated using the readDNV function. Returns a DNV object containing 
only INDELs.
```{r}
indels <- extractINDELs(input)
nrow(indels)
```

### Extract MNVs Only (There are None in the Test Set)
extractMNVs = Extracts multi-nucleotide variants (MNVs) out from a DNV object
generated using the readDNV function. Returns a DNV object containing only MNVs.
```{r}
denovoMNVs <- readDNV(paste(path.package("acorn"),"/extdata/mnv_test.txt",sep="")) 

mnvs <- extractMNVs(denovoMNVs)
nrow(mnvs)
```

### Calculate the MNV Lengths
calculateMNVLengths = This function will automatically grab only 
the MNVs from the DNV object for the calculation of the MNV lengths 
ratio. Returns the length of the MNVs, in the form of an object, 
observed in the DNV object. It also returns a barplot of the MNV lengths.
```{r}
calculateMNVLengths(mnvs)
```

### Calculate the Transition/Transversion Ratio
calculateTiTvRatio = This function will automatically grab only the SNVs from 
the DNV object for the calculation of the transition/transversion (Ti/Tv) ratio.
Returns the counts of transitions, the counts of transversions, the Ti/Tv ratio,
and a barplot of the different types of SNV changes observed in the DNV object.
```{r}
calculateTiTvRatio(input)
```

### Calculate the Deletion/Insertion Ratio
calculateDeletionInsertionRatio = This function will automatically grab only 
the INDELs from the DNV object for the calculation of the deletion/insertion 
ratio. Returns the counts of deletions, the counts of insertions, and the
deletion/insertion ratio.
```{r}
calculateDeletionInsertionRatio(input)
```

### Calculate Deletion Lengths
calculateDeletionLengths = This function will automatically grab only the 
deletions from the DNV object for the calculation of the length of the 
deletions. Returns the length of the deletions, in the form of an object, 
observed in the DNV object. It also returns a barplot of the deletion lengths.
```{r}
dellengths <- calculateDeletionLengths(input)
head(dellengths)
```

### Calculate Insertion Lengths
calculateInsertionLengths = This function will automatically grab only the 
insertions from the DNV object for the calculation of the length of the 
insertions. Returns the length of the insertions, in the form of an object, 
observed in the DNV object. It also returns a barplot of the insertion lengths.
```{r}
inslengths <- calculateInsertionLengths(input)
head(inslengths)
```

### Keep Only the Autosomes
extractAutosomes = Extracts the autosomes (chromosomes 1 to 22) out from a DNV 
object originally generated using the readDNV function. You can also run this 
on objects generated from extractSNVs, extractINDELs, or extractMNVs. Returns 
a DNV object containing only DNVs on the autosomes.
```{r}
aut <- extractAutosomes(input)
nrow(aut)
table(aut[,2])
```

### Keep Only the X Chromosome
extractX = Extracts the X chromosome out from a DNV object originally generated 
using the readDNV function. You can also run this on objects generated from
extractSNVs, extractINDELs, or extractMNVs. Returns a DNV object containing 
only DNVs on the X chromosome.
```{r}
X <- extractX(input)
nrow(X)
table(X[,2])
```

### Keep Only the Y Chromosome (There are Mone on the Y in the Test Dataset)
extractY = Extracts the Y chromosome DNVs out from a DNV object originally 
generated using the readDNV function. You can also run this on objects 
generated from extractSNVs, extractINDELs, or extractMNVs. Returns a DNV 
object containing only DNVs on the Y chromosome.
```{r}
Y <- extractY(input)
nrow(Y)
```

### Calculate Counts per Individual
countsPerIndividual = This function will count the DNVs from a DNV object 
originally generated using the readDNV function. You can also run this on 
objects generated from extractSNVs, extractINDELs, or extractMNVs. Returns 
the mean of the DNV counts per individual, the standard deviation of the DNV 
counts per individual, a plot of the density of the DNV counts per individual,
and an object consisting of the sample name and the counts of their DNVs that
can be assigned to another object.
```{r}
counts <- countsPerIndividual(input)
head(counts)
```

### Load in Example Data for Parental Age Analyses
```{r}
input <- readDNV(paste(path.package("acorn"),"/extdata/dnms_from_Ng_et_al_2022_Human_Mutation_paper.txt.gz",sep="")) 


countExample <- read.delim(paste(path.package("acorn"),"/extdata/dnm_count_example.txt",sep=""))
parentExample <- read.delim(paste(path.package("acorn"),"/extdata/parental_age_example.txt",sep=""))
```

### Make Parental Age Object
parentalAgeObject = Takes in a counts object that is either the result of
countsPerIndividual() or is already read into an object from a file that 
contains the following two fields: sample and number of DNVs. The parental 
age object should be read in and contain the following fields: sample, 
father age at child's birth, and mother age at child's birth. Returns back
an object with the de novo counts and parental age data together. The 
fields in this file are sample, dnm_counts, fatherAge, and motherAge.
```{r}
parents <- parentalAgeObject(countExample, parentExample)
```

### Run Parental Age Analyses Including Both Mother and Father
parentalAge = This function will calculate the correlation between father's 
and mother's age at birth and DNV counts per individual, the results of the 
linear model taking the form: lm(formula = dnm_counts ~ fatherAge+motherAge, 
data = parentalAgeObject). Input required is output from the 
parentalAgeObject function in this package. Returns the results of the 
linear model taking the form: lm(formula = dnm_counts ~ fatherAge + motherAge, 
data = parentalAgeObject). It also returns a plot of father's and mother's age 
at birth and DNV counts.
```{r}
parentalAge(parents)
```

### Run Parental Age Analyses for Father Age Only
fatherAge = This function will calculate the correlation between father's 
age at birth and DNV counts per individual, the results of the linear model 
taking the form: lm(formula = dnm_counts ~ fatherAge, data = parentalAgeObject).
Input required is output from the parentalAgeObject function in this package.
Returns the correlation between father's age at birth and DNV counts per 
individual and the results of the linear model taking the form: lm(formula =
dnm_counts ~ fatherAge, data = parentalAgeObject). It also returns a plot 
of father's age at birth and DNV counts.
```{r}
fatherAge(parents)
```

### Run Parental Age Analyses for Mother Age Only
motherAge = This function will calculate the correlation between mother's 
age at birth and DNV counts per individual, the results of the linear model 
taking the form: lm(formula = dnm_counts ~ motherAge, data = parentalAgeObject).
Input required is output from the parentalAgeObject function in this package.
Returns the correlation between mother's age at birth and DNV counts per 
individual and the results of the linear model taking the form: lm(formula =
dnm_counts ~ motherAge, data = parentalAgeObject). It also returns a plot of
mother's age at birth and DNV counts.
```{r}
motherAge(parents)
```

### Print the Session Information
```{r}
sessionInfo()
```
