---
title: "acorn Vignette"
author: "Tychele N. Turner, Ph.D., Washington University School of Medicine"
date: "2023-01-18"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{acorn Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### call up acorn
```{r}
library('acorn')
```

### check out acorn functions
```{r}
lsf.str("package:acorn")
```

### load in test data
```{r}
input <- readDNV(paste(path.package("acorn"),"/extdata/dnms_from_Ng_et_al_2022_Human_Mutation_paper.txt.gz",sep="")) 
head(input)
str(input)
```

### extract individual
```{r}
ind <- extractIndividual(input, "HG01928")
head(ind)
nrow(ind)
table(ind[,1])
```

### extract individuals if you want
```{r}
ind <- extractIndividual(input, c("HG01928", "HG03915"))
head(ind)
nrow(ind)
table(ind[,1])
```

### extract SNVs
```{r}
snvs <- extractSNVs(input)
nrow(snvs)
```

### extract indels
```{r}
indels <- extractINDELs(input)
nrow(indels)
```

### extract MNVs (there are none in the test set)
```{r}
mnvs <- extractMNVs(input)
nrow(mnvs)
```

### calculate the Transition/Transversion ratio
```{r}
calculateTiTvratio(input)
```

### calculate deletion/insertion ratio
```{r}
calculateDeletionInsertionratio(input)
```

### calculate deletion lengths
```{r}
dellengths <- calculateDeletionLengths(input)
head(dellengths)
```

### calculate insertion lengths
```{r}
inslengths <- calculateInsertionLengths(input)
head(inslengths)
```

### keep only the autosomes
```{r}
aut <- extractAutosomes(input)
nrow(aut)
table(aut[,2])
```

### keep only the X chromosome
```{r}
X <- extractX(input)
nrow(X)
table(X[,2])
```

### keep only the Y chromosome (there are none on the Y in the test dataset)
```{r}
Y <- extractY(input)
nrow(Y)
```

### calculate counts per individual
```{r}
counts <- countsPerIndividual(input)
head(counts)
```

### load in example data for parental age analyses
```{r}
input <- readDNV(paste(path.package("acorn"),"/extdata/dnms_from_Ng_et_al_2022_Human_Mutation_paper.txt.gz",sep="")) 


countExample <- read.delim(paste(path.package("acorn"),"/extdata/dnm_count_example.txt",sep=""))
parentExample <- read.delim(paste(path.package("acorn"),"/extdata/parental_age_example.txt",sep=""))
```

### make parental age object
```{r}
parents <- parentalAgeObject(countExample, parentExample)
```

### run parental age analyses including both mother and father
```{r}
parentalAge(parents)
```

### run parental age analyses for father age only
```{r}
fatherAge(parents)
```

### run parental age analyses for mother age only
```{r}
motherAge(parents)
```





